---
title: "Analysis of PJM Interconnection Queue Data"
subtitle: "Annika Barth PS4, Spring 2025"
format: html
editor: visual
echo: false
message: false
warning: false
---

```{r, echo=FALSE}
# Set Up

library(magrittr)
library(dplyr)
library(tidyverse)
library(readxl)
library(httr)
library(lubridate)
library(ggplot2)
library(gt)

# Read in PJMCycleProjects file

url <- "https://github.com/abarth15/PDM-Final-Project/raw/main/PJMCycleProjects.xlsx"

file <- tempfile(fileext = ".xlsx")
download.file(url, file, mode = "wb")

pjm_full_cycle <- read_excel(file)

# Read in PJMActive Projects file

url <- "https://github.com/abarth15/PDM-Final-Project/raw/main/PJMActiveProjects.xlsx"

file <- tempfile(fileext = ".xlsx")
download.file(url, file, mode = "wb")

pjm_in_service <- read_excel(file)
```

## Introduction and Data Description

This analysis explores PJM's interconnection queue data, which includes characteristics of generators seeking to connect to the grid. This information can help us understand what the resource mix of the future might look like and which communities are most likely to come into increasing contact with energy development. It can also help us evaluate the success of grid infrastructure planning and whether planning practices are meeting the needs of future generators.

PJM, the regional grid operator in the Mid-Atlantic and the nation's largest grid operator, houses interconnection data in online, publicly available, searchable databases. This analysis makes use of two data sets from these resources: (1) generators that are connected to the grid and in service; and (2) generators that are active in PJM's interconnection queue, seeking to interconnect the grid, but have not yet received permission to do so. The first data set was pulled from PJM's serial service request status database, which catalogues generators that are either in service or moving through various stages of the interconnection process. PJM recently instituted a new method of studying interconnection requests, and projects participating in this process are stored in PJM's cycle service request status database. The second datas et was pulled from this database. In both data sets, each electric generator or power plant is the unit of observation.

For this analysis, both data sets were filtered for generation interconnection project type before download. The serial service request status data set was filtered for only projects that are in service. The filtered data set contains 1,204 observations and 43 variables, while the cycle service request status data set, which was filtered for only projects that are active in the queue, contains 862 observations and 41 variables. For simplicity, projects seeking interconnection in PJM's serial service request status database were not included in this analysis. The most important variables in these data sets are the fuel type, MW of energy, the state and county where the generator is located, the submitted date, requested in-service date, and actual in-service date.

## Data Analysis

**Central variables:**

-   **Fuel:** This variable describes the fuel used to generate electricity. Some examples include coal, oil, natural gas, solar, wind, and biomass. This variable is central to the missions of climate and clean energy advocacy organizations and to the policy goals of states in the PJM region that have renewable portfolio standards. The fuel type of generators in the interconnection queue is important to any government or civic entity seeking to understand the future of the electric power sector and to prepare supply chains, infrastructure, and the workforce to align with future power generation.

-   **MW Energy:** This variable describes the maximum amount of power the generator is capable of injecting into the grid.

-   **State:** The state where the generator will be built. This can help clean energy nonprofits determine where to target advocacy and outreach efforts.

-   **County:** The county where the generator will be built. These are the locations that will see energy projects developed in their communities. Outreach and engagement is critical in these locations because most permitting decisions happen at the local level, and community opposition can quickly kill a project.

-   **Submitted Date:** The date the project developer submitted an interconnection request to PJM. This variable is critical to determining how long it takes a generator to move through the interconnection queue.

-   **Actual In Service Date:** The date the generator came online and began providing power to the grid. This is also important to determining how long it takes a generator to move through the interconnection queue. This variable is, of course, only available for generators that are actually in service and are no longer waiting in the interconnection queue.

### Fuel Types in the Queue

The two tables below show: (a) fuel types as a percent of the total number of projects in PJM's interconnection queue; and (b) fuel types as a percent of the total MW in the queue. Solar projects account for a majority of projects in the queue, at almost 60%, followed by storage, hybrid projects (solar paired with battery storage), and wind. While offshore wind and natural gas both account for less than 1% of the number of projects in the queue, they account for roughly 5% each of the MW in the queue, indicating that there are a handful of large projects of these fuel types. Conversely, while solar generators account for almost 60% of the projects in the queue, they account for a slightly lower percent of total MW, at around 50%. Still, this analysis demonstrates that solar and storage make up the decisive majority of the electric generating units and the MW that will come online to serve electricity consumers in the PJM region in the near- and medium- term. These are the projects that the market is procuring, so these are the projects that policymakers, grid infrastructure, supply chains, and the workforce should be prepared to accommodate.

```{r}

fuel_percent_projects <- filter(pjm_full_cycle, Status == "Active") %>%
  group_by(Fuel) %>%
  summarise(count = n()) %>%
  mutate(`Percent of Projects` = round(100 * count / sum(count), 1)) %>%
  select(Fuel, `Percent of Projects`) %>%
  arrange(desc(`Percent of Projects`))

gt(fuel_percent_projects) %>%
  tab_header(title = "Fuel Type as Percent of Total Projects")
```

```{r}

pjm_full_cycle$`MW Energy` <- as.numeric(pjm_full_cycle$`MW Energy`)

fuel_percent_energy <- filter(pjm_full_cycle, Status == "Active") %>%
  group_by(Fuel) %>%
  summarize(TotalMW = sum(`MW Energy`, na.rm = TRUE)) %>%
  mutate(`Percent of MW` = round((TotalMW / sum(TotalMW)) * 100, 1)) %>%
  select(Fuel, `Percent of MW`) %>%
  arrange(desc(`Percent of MW`))

gt(fuel_percent_energy) %>%
  tab_header(title = "Fuel Type as Percent of Total MW")
```

### States and Counties with the Most Clean Energy in the Queue

The tables below show the top locations where developers are seeking to build clean energy projects and connect them to the grid. This tells us where projects are likely to be built in the future, not necessarily where they are already located. Clean energy in this analysis is defined as wind, solar, storage, and hybrid projects. As shown below, Virginia, Illinois, and Indiana are expecting the most clean energy development of the 13 states in the PJM region. The top five counties in PJM for future clean energy development are all in Illinois and Virginia. This means that these states should be priority locations for clean energy advocacy organizations to focus their efforts. State policymakers should be made aware of the amount of clean energy that developers are seeking to build in their states, and the economic growth opportunities associated with that development should those projects be allowed to come online.

```{r}

top_three_states <- filter(pjm_full_cycle, Status == "Active" & Fuel %in% 
    c("Solar", "Wind", "Storage", "Solar,Storage,Hybrid")) %>% 
  group_by(State) %>%
  summarize(TotalMW = sum(`MW Energy`, na.rm = TRUE)) %>%
  mutate(`Clean Energy MW` = TotalMW) %>%
  select(State, `Clean Energy MW`) %>%
  arrange(desc(`Clean Energy MW`)) %>%
  head(3)

gt(top_three_states) %>%
  tab_header(title = "Top Three States for Clean Energy in PJM's Queue")

```

The clean energy sector should prioritize supporting robust community engagement efforts in Lee, McHenry, and McLean counties in Illinois, and Brunswick and Halifax counties in Virginia, since these counties are where most clean energy developers will seek to build their projects in the future. Community engagement well ahead of project development is key to building public trust and acceptance of clean energy projects. Besides transmission capacity constraints, public opposition and local siting challenges are the biggest barriers to deploying more clean energy. Community engagement is a critical solution to those challenges, and it is most effective when it happens early, often, and well in advance of project construction, which is why it is important to prioritize community engagement in locations that are likely to see an increase in development in the future - not necessarily in locations where an abundance of projects already exists.

```{r}

top_five_counties <- filter(pjm_full_cycle, Status == "Active" & Fuel %in% 
    c("Solar", "Wind", "Storage", "Solar,Storage,Hybrid")) %>% 
  group_by(State, County) %>%
  summarize(TotalMW = sum(`MW Energy`, na.rm = TRUE)) %>%
  mutate(`Clean Energy MW` = TotalMW) %>%
  select(State, County, `Clean Energy MW`) %>%
  arrange(desc(`Clean Energy MW`)) %>%
  head(5)

gt(top_five_counties) %>%
  tab_header(title = "Tope Five Counties for Clean Energy in PJM's Queue")
```

### Interconnection Wait Times

The chart below shows the average amount of time that elapsed between the date a project submitted an interconnection request and the in-service date for that project. This is an approximate although imperfect measure for the amount of time projects spend waiting in PJM's interconnection queue. A more accurate measure would be the amount of time between submitting a request and receiving an interconnection agreement, but that information was not available in the dataset. Projects that receive an interconnection agreement might not ultimately be completed for a variety of reasons, and projects can still be subject to permitting and other delays that affect the in-service date even after receiving an interconnection agreement from PJM.

Nevertheless, the chart below is still indicative of the amount of time generators must wait before connecting to the grid after submitting an interconnection request. As shown, the amount of time has increased over the last two decades and varies by fuel type. Solar and storage projects that came online in the early 2010's did so less than 2.5 years after submitting an interconnection request - much faster than natural gas plants, which at times waited almost 4 years. But by the early 2020's, solar, storage, and natural gas plants that began providing service in those years all waited, on average, more than 5 years after submitting an interconnection request. The average time between in-service date and interconnection request submission across all fuel types was just under 2.5 years in 2010 and now sits at around 7 years, a decade and a half later.

```{r}

time_in_queue <- pjm_in_service %>%
  mutate(across(c(`Submitted Date`, `Actual In Service Date`), ~ 
    as_date(mdy(.x)))) %>%
  mutate(`Time In Queue` = interval(`Submitted Date`, `Actual In Service Date`)
    / years(1))

mean_time_fuel <- time_in_queue %>%
  mutate(`In Service Year` = year(`Actual In Service Date`)) %>%
  group_by(Fuel, `In Service Year`) %>%
  summarize(AverageTime = mean(`Time In Queue`, na.rm = TRUE)) %>%
  select(Fuel, `In Service Year`, AverageTime) %>%
  filter(`In Service Year` >= 2005 & Fuel %in% c("Solar", "Wind", "Storage", 
    "Natural Gas")) %>%
  rename(, `Average Time In Queue` = AverageTime)

all_fuels <- mean_time_fuel %>%
  group_by(`In Service Year`) %>%
  summarize(`Average Time In Queue` = mean(`Average Time In Queue`,
    na.rm = TRUE)) %>%
  mutate(Fuel = "All Fuels")

mean_time_all_fuel <- bind_rows(mean_time_fuel, all_fuels)

ggplot(mean_time_all_fuel, aes(`In Service Year`, `Average Time In Queue`, 
  color = Fuel)) + geom_line() +
  scale_color_manual(values = c(
    "Solar" = "yellow", 
    "Wind" = "skyblue",
    "Storage" = "green",
    "Natural Gas" = "grey",
    "All Fuels" = "black")) +
  labs(title = "Interconnection Wait Times by Fuel and Year")

```

The chart below shows the z-score for average wait times in the interconnection queue by fuel type relative to the average wait time across all fuels for each year between 2005 and 2025. Based on this chart, it is clear that wind projects face wait times much higher than the all-fuel average wait time and that, in recent years, solar has tracked closely with the all-fuel average. Storage projects consistently move through the interconnection queue faster than the yearly average across all fuels. Solar, storage, and natural gas projects are likely to come online in the most timely manner, so policy makers concerned with meeting new electricity demand quickly should focus on these resources. Advocates and industry analysts could consider investigating factors that impede wind projects from moving through the interconnection queue.

```{r}

sd_time_in_queue <- time_in_queue %>%
  mutate(`In Service Year` = year(`Actual In Service Date`)) %>%
  filter(`In Service Year` >= 2005 & Fuel %in% c("Solar", "Wind", "Storage", 
    "Natural Gas")) %>%
  group_by(`In Service Year`) %>%
  summarize(
    `All Fuel Mean` = mean(`Time In Queue`, na.rm = TRUE),
    `All Fuel SD` = sd(`Time In Queue`, na.rm = TRUE)
  )

z_score <- mean_time_fuel %>%
  left_join(sd_time_in_queue, by = "In Service Year") %>%
  mutate(
    `Z-Score` = (`Average Time In Queue` - `All Fuel Mean`) / `All Fuel SD`
  )

ggplot(z_score, aes(x = `In Service Year`, y = `Z-Score`, color = Fuel)) +
  geom_line() +
  geom_point() +
  scale_color_manual(values = c(
    "Solar" = "yellow", 
    "Wind" = "skyblue",
    "Storage" = "green",
    "Natural Gas" = "grey"))+
  labs(title = "Z-Score of Average Interconnection Queue Time by Fuel Type")
```

The drivers behind increasing wait times in the interconnection queue and the implications for the power sector and electricity consumers are beyond the scope of this analysis. Hopefully, this analysis will equip clean energy advocacy organiztions and other stakeholders with basic information to demonstrate to relevant decision makers important trends in generator interconnection in PJM and make the case for needed changes to support a clean, affordable, and reliable power system.

## Data Citations

Link to GitHub repository: <https://github.com/abarth15/PDM-Final-Project>

PJM. (2025). *Cycle Service Request Status* \[Data set\]. PJM. <https://www.pjm.com/planning/m/cycle-service-request-status>

PJM. (2025). *Serial Service Request Status* \[Data set\]. PJM. <https://www.pjm.com/planning/service-requests/serial-service-request-status>
